---
title: "Homework 2"
author: "Zhiquan Chen"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(janitor)
library(readxl)
```

# Problem 1
```{r}
subway_raw =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      .default = col_character(),
      `Station Latitude` = col_double(),
      `Station Longitude` = col_double(),
      ADA = col_logical()
    )
  ) |>
clean_names()
subway_df =
  subway_raw |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    starts_with("route"),
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = entry == "YES"
  )

```
This dataset records subway entrances and exits in NYC. After cleaning, it contains `r dim(subway_df)[1]` rows and `r dim(subway_df)[2]` columns, including line, station name, station coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. I standardized the column names, retained the requested variables, explicitly specified data types, and converted entry from “YES”/“NO” to TRUE/FALSE. No entrance/exit records were removed. The data are not yet tidy for route-level analysis because route names are stored in eleven separate columns, route1 through route11, instead of a single route-name column.

```{r}
n_stations =
  subway_df |>
  distinct(line, station_name) |>
  nrow()
n_stations
```
There are `r n_stations` distinct stations.

```{r}
n_ada_stations =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
n_ada_stations
```
`r n_ada_stations` stations are ADA compliant.

```{r}
no_vending_summary =
  subway_df |>
  filter(vending == "NO") |>
  summarise(
    n_no_vending = n(),
    n_allow_entry = sum(entry),
    prop_allow_entry = mean(entry)
  )

no_vending_summary
```

`r sprintf("%.2f%%", no_vending_summary$prop_allow_entry * 100)` of station entrances / exits without vending allow entrance.

```{r}
subway_long =
  subway_df |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    names_transform = list(route_number = as.integer),
    values_to = "route_name",
    values_drop_na = TRUE
  )
```

```{r}
n_a_stations =
  subway_long |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

n_a_stations
```

`r n_a_stations` distinct stations serve the A train.

```{r}
n_a_ada_stations =
  subway_long |>
  filter(
    route_name == "A",
    ada == TRUE
  ) |>
  distinct(line, station_name) |>
  nrow()

n_a_ada_stations
```
`r n_a_ada_stations` of the stations that serve the A train are ADA compliant.

# Problem 2

```{r}
trash_path = "data/202509 Trash Wheel Collection Data.xlsx"

mr_trash_df =
  read_excel(
    trash_path,
    sheet = "Mr. Trash Wheel",
    range = "A2:N710"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    month = str_to_title(str_trim(month)),
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )

professor_trash_df =
  read_excel(
    trash_path,
    sheet = "Professor Trash Wheel",
    range = "A2:M135"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    month = str_to_title(str_trim(month)),
    year = as.integer(year),
    date = as.Date(date),
    trash_wheel = "Professor Trash Wheel"
  )

gwynnda_trash_df =
  read_excel(
    trash_path,
    sheet = "Gwynns Falls Trash Wheel",
    range = "A2:L352"
  ) |>
  clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    dumpster = as.integer(dumpster),
    month = str_to_title(str_trim(month)),
    year = as.integer(year),
    date = as.Date(date),
    trash_wheel = "Gwynnda"
  )

trash_df =
  bind_rows(
    mr_trash_df,
    professor_trash_df,
    gwynnda_trash_df
  )
```

```{r}
professor_summary =
  trash_df |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarise(
    total_weight_tons = sum(weight_tons, na.rm = TRUE)
  )

professor_summary$total_weight_tons
```
```{r}
gwynnda_june_2022_summary =
  trash_df |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarise(
    n_records = n(),
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )

gwynnda_june_2022_summary
```

The combined dataset contains `r nrow(trash_df)` dumpster-level observations
and `r ncol(trash_df)` variables: `r nrow(mr_trash_df)` observations from
Mr. Trash Wheel, `r nrow(professor_trash_df)` from Professor Trash Wheel,
and `r nrow(gwynnda_trash_df)` from Gwynnda. Key variables include the
source trash wheel, dumpster identifier, month, year, collection date,
weight in tons, volume in cubic yards, and recorded counts of items
such as plastic bottles, cigarette butts, and sports balls. I restricted
each import to the relevant table range, standardized variable names
and month labels, removed rows without dumpster identifiers, made
year and date types consistent, and rounded Mr. Trash Wheel's sports-ball
counts to the nearest integer before converting them to integer type.
A source variable identifies the trash wheel for each observation.
Missing measurements and variables absent from individual sheets are
retained as NA, not zero. For the available data, Professor Trash Wheel
collected `r sprintf("%.2f", professor_summary$total_weight_tons)` tons of
trash, and Gwynnda collected
`r format(gwynnda_june_2022_summary$total_cigarette_butts, big.mark = ",", scientific = FALSE, trim = TRUE)`
cigarette butts in June 2022.

