---
title: "Homework 2"
author: "Zhiquan Chen"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(janitor)
```

# Problem 1
```{r}
subway_raw =
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      .default = col_character(),
      `Station Latitude` = col_double(),
      `Station Longitude` = col_double(),
      ADA = col_logical()
    )
  ) |>
clean_names()
subway_df =
  subway_raw |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    starts_with("route"),
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = entry == "YES"
  )

```
This dataset records subway entrances and exits in NYC. After cleaning, it contains `r dim(subway_df)[1]` rows and `r dim(subway_df)[2]` columns, including line, station name, station coordinates, routes served, entry permission, vending availability, entrance type, and ADA compliance. I standardized the column names, retained the requested variables, explicitly specified data types, and converted entry from “YES”/“NO” to TRUE/FALSE. No entrance/exit records were removed. The data are not yet tidy for route-level analysis because route names are stored in eleven separate columns, route1 through route11, instead of a single route-name column.

```{r}
n_stations =
  subway_df |>
  distinct(line, station_name) |>
  nrow()
```
There are `r n_stations` distinct stations.

```{r}
n_ada_stations =
  subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```
`r n_ada_stations` stations are ADA compliant.

```{r}
no_vending_summary =
  subway_df |>
  filter(vending == "NO") |>
  summarise(
    n_no_vending = n(),
    n_allow_entry = sum(entry),
    prop_allow_entry = mean(entry)
  )

no_vending_summary
```

`r no_vending_summary[3]` of station entrances / exits without vending allow entrance.